How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2023-05-09 analysis/paper/paper.Rmd
updated text

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca. 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human]. Stemmed points are projectile points made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft. If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea. However, most open sites in Korea do not preserve organic remains due to acidic soil, so shafts have not been found. There are very few limestone cave sites containing human fossils or faunal remains (e.g. Durubong and Jeommal Cave), which located in constrained area to access in North Korea or have been rarely analyzed to directly apply to other archaeological research[@bae2013paleolithic]. Therefore stone artifacts such as stemmed points are our only window to examine hunting and subsistence strategies. Here we apply a behavioral ecological framework to explore questions about the implications of the appearance of this new technology. We use data from 23 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.

>>>>

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (ca. 42-35 ka) in Korea is often assumed to have transformed forager lifeways because it was thought to have indicated more specialized hunting practices [@seong2008tanged; @chang2013human]. Stemmed points are projectile points made from an elongated flake or blade with slight retouch on the distal end to shape an acute tip and on the proximal end to make a stem, which connects to a shaft. These types of artifacts are often called tanged points in the western hemisphere [@serwatka20162d; @ivanovaite2020all], but we use ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent a hafting projection used for joining the lithic to a wooden shaft. If these artifacts were projectile weapon tips, then stemmed points represent the first appearance of hafting technology in Korea. However, most open sites in Korea do not preserve organic remains due to acidic soil, so shafts have not been found. There are very few limestone cave sites containing human fossils or faunal remains (e.g. Durubong and Jeommal Cave), which located in constrained area to access in North Korea or have been rarely analyzed to directly apply to other archaeological research[@bae2013paleolithic]. Therefore stone artifacts such as stemmed points are our only window to examine hunting and subsistence strategies. Here we apply a behavioral ecological framework to explore questions about the implications of the appearance of this new technology. We use data from 23 sites in South Korea to explore changes in the way people occupied the landscape, the way people used sites, and the social and ecological contexts associated with the appearance stemmed points, and the blade industry that followed.
